POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2023-10-11 challenges/linearity.tex
inserted dual calculus in linearity section and other non-insignificant changes

\subsection{Challenge: Linearity and behavioral type systems}
\label{sec:challenge:linearity-beh-types}

This challenge formalises a proof that requires reasoning about linearity of channels.
Linearity is the notion that a channel must be used exactly once in a process.
This is necessary to prove properties about session type systems.
Linear reasoning is also necessary to formalise e.g.\ linear and affine types for the pi-calculus and cut elimination in linear logics.

The setting for this challenge is a small calculus with a session type system, the syntax and semantics of which are presented below.
The objective of this challenge is to prove type preservation (also known as subject reduction) for the type system, i.e.\ that well-typed processes can only transition to processes which are also well-typed in the same context.

@@

\begin{metanote}
  AM: I've moved the discussion of well-formedness must go after the
  type system, since we want to separate the preservation vs
  safety challenges
\end{metanote}
\paragraph{Syntax}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \) and \emph{type contexts} \( \Gamma \) is as follows:
\begin{metanote}
  AM: take the syntax from dual file appst.tex and adapt
\end{metanote}
\begin{align*}
  S,T := &&& \Tend \\
  |&&& \Tbase \\
  |&&& \Tin{S} \\
  |&&& \Tout{S} \\
  \Gamma := &&& \Cempty \\
  |&&& \Cadd{\Gamma}{\hastype{x}{S}}
       %% AM: added
        \\
  |&&& \Cadd{\Gamma}{\hastype{l}{\Tbase}} %% changed, AM
\end{align*}
The \emph{end type}, \( \Tend \), is the type of an endpoint on which no further interaction is possible.
The \emph{base type}, \( \Tbase \), is the type of base values.
The \emph{input type}, \( \Tin{S} \), is the type of an endpoint which is ready to receive a value, then continue with type \( S \).
The \emph{output type}, \( \Tout{S} \), is the type of an endpoint which is ready to send a value, then continue with type \( S \).

Typing contexts gather type information about names.
The \emph{empty type context}, \( \Cempty \), does not contain any information.
The \emph{type context with \( x \) added}, \( \Cadd{\Gamma}{\hastype{x}{T}} \) adds the information that \( x \) has type \( T \) to the existing type context \( \Gamma \).
All of the names added to a type context must be distinct.
Note that
the order in which information is added to a type context thus does not matter.
% We treat type contexts as sets, not ordered lists, and the order in which information is added to a type context thus does not matter.

@@

\begin{metanote}
  AM: put here the dual rules, remove splitting
\end{metanote}

To keep track of linearity, we introduce two \emph{qualifiers}: \( \un{T} \) and \( \lin{T} \).
A type is \emph{unrestricted}, \( \un{T} \), if and only if \( T = \Tbase \) or \( T = \Tend \).
All types are \emph{linear}, \( \lin{T} \).
The qualifiers extend to contexts in the following way: \( q(\Gamma) \) if and only if \( \hastype{x}{T} \in \Gamma \) implies \( q(T) \).

The \emph{linear split} of a context, \( \Gamma = \Csplit{\Gamma_1}{\Gamma_2} \), is defined by the following rules:
\begin{mathpar}
  \inferrule[Split-Empty]{ }{\Cempty = \Csplit{\Cempty}{\Cempty}}
  \and
  \inferrule[Split-Un]{\Csplit{\Gamma_1}{\Gamma_2} = \Gamma \\ \un{T}}{\Cadd{\Gamma}{\hastype{x}{T}} = \Csplit{(\Cadd{\Gamma_1}{\hastype{x}{T}})}{(\Cadd{\Gamma_2}{\hastype{x}{T}})}}
  \and
  \inferrule[Split-Lin-L]{\Gamma = \Csplit{\Gamma_1}{\Gamma_2}}{\Cadd{\Gamma}{\hastype{x}{S}} = \Csplit{(\Cadd{\Gamma_1}{\hastype{x}{S}})}{\Gamma_2}}
  \and
  \inferrule[Split-Lin-R]{\Gamma = \Csplit{\Gamma_2}{\Gamma_2}}{\Cadd{\Gamma}{\hastype{x}{S}} = \Csplit{\Gamma_1}{(\Cadd{\Gamma_2}{\hastype{x}{S}})}}
\end{mathpar}

@@

We have three typing judgments: one for values, one for names, and one for processes.
The typing rules for values are as follows:
\begin{mathpar}
  \inferrule[T-Base]{\un{\Gamma}}{\types{\Gamma}{\hastype{a}{\Tbase}}}
  \and
  \inferrule[T-Var]{\un{\Gamma}}{\types{\Cadd{\Gamma}{\hastype{l}{\Tbase}}}{\hastype{l}{\Tbase}}}
\end{mathpar}
The only typing rule for names is:
\begin{mathpar}
  \inferrule[T-Name]{\un{\Gamma}}{\types{\Cadd{\Gamma}{\hastype{x}{T}}}{\hastype{x}{T}}}
\end{mathpar}
The typing rules for processes are as follows:
\begin{mathpar}
  \inferrule[T-Inact]{\un{\Gamma}}{\types{\Gamma}{\Pend}}
  \and
  \inferrule[T-Par]{\types{\Gamma_1}{P} \\ \types{\Gamma_2}{Q}}{\types{\Csplit{\Gamma_1}{\Gamma_2}}{\Ppar{P}{Q}}}
  \and
  \inferrule[T-Res]{\types{\Cadd{\Cadd{\Gamma}{\hastype{x}{T}}}{\hastype{y}{\dual{T}}}}{P}}{\types{\Gamma}{\Presd{x}{y}{P}}}
  \and
  \inferrule[T-Out]{\types{\Gamma_1}{\hastype{x}{\Tout{T}}} \\ \types{\Gamma_2}{\hastype{v}{\Tbase}} \\ \types{\Cupdate{\Gamma_3}{\hastype{x}{T}}}{P}}{\types{\Csplit{\Gamma_1}{\Csplit{\Gamma_2}{\Gamma_3}}}{\Pout{x}{v}{P}}}
  \and
  \inferrule[T-In]{\types{\Gamma_1}{\hastype{x}{\Tin{U}}} \\ \types{\Cupdate{\Cadd{\Gamma_2}{\hastype{l}{\Tbase}}}{\hastype{x}{U}}}{P}}{\types{\Csplit{\Gamma_1}{\Gamma_2}}{\Pin{x}{l}{P}}}
\end{mathpar}

\subsubsection{Challenge}
\begin{metanote}
AM:  I'm separating sr from safety, or whatever we call the latter
\end{metanote}
The objective of this challenge is to prove type preservation for our calculus with session types. We start with:
\begin{theorem}[Subject reduction]
  If \( \types{\Gamma}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma}{Q} \).
\end{theorem}

The following lemmas may be of use:
\begin{lemma}
  If \( \types{\Gamma}{P} \) and \( \un{T} \) then \( \types{\Cadd{\Gamma}{\hastype{x}{T}}}{P} \).
\end{lemma}

\begin{lemma}
  Assume \( \types{\Gamma}{P} \) and \( x \notin \freenames{P} \).
  Then:
  \begin{enumerate}
  \item \( \hastype{x}{\Tout{T}} \notin \Gamma \).
  \item \( \hastype{x}{\Tin{T}} \notin \Gamma \).
  \item if \( \Gamma = \Cadd{\Gamma'}{\hastype{x}{T}} \) then \( \types{\Gamma'}{P} \).
  \end{enumerate}
\end{lemma}

\begin{lemma}
  If \( \types{\Gamma}{P} \) and \( \scong{P}{Q} \) then \( \types{\Gamma}{Q} \).
\end{lemma}
% \begin{metanote}
%   AM: the statement of substitution was wrong: now it is consistent with the op sem, but I need to check the (I assume now simpler) proof. Note that $T$ must be \textbf{base} and \( \un{\Gamma_i} \) must also hold.
% \end{metanote}
\begin{lemma}
  If \( \types{\Gamma_1}{\hastype{v}{T}} \) and
  \( \types{\Cadd{\Gamma_2}{\hastype{l}{T}}}{P} \) and
  \( \Gamma = \Csplit{\Gamma_1}{\Gamma_2} \) then
  \( \types{\Gamma}{\subst{P}{l}{v}} \).
\end{lemma}

>>>>

\subsection{Challenge: Linearity and behavioral type systems}
\label{sec:challenge:linearity-beh-types}
\renewcommand{\Cempty}{\cdot}
\renewcommand{\Csplit}[2]{#1,  #2}
\renewcommand{\Cupdate}[2]{#1, #2}
\newcommand{\TinV}[2]{{?#1}.#2}
\newcommand{\ToutV}[2]{{!#1}.#2}
\newcommand{\tend}[1]{\mathbf{end}(#1)}
This challenge formalises a proof that requires reasoning about linearity of channels.
Linearity is the notion that a channel must be used exactly once in a process.
This is necessary to prove properties about session type systems.
Linear reasoning is also necessary to formalise e.g.\ linear and affine types for the pi-calculus and cut elimination in linear logics.

The setting for this challenge is a small calculus with a session type
system, the syntax and semantics of which are given below. The
calculus is a fragment of the one presented in~\cite{Vasconcelos2012},
formulated in the dual style of~\cite{barber96tr}.

The objective of this challenge is to prove type preservation (also
known as subject reduction) for the type system, i.e.\ that well-typed
processes can only transition to processes which are also well-typed
in the same context.

@@

\begin{metanote}
  AM: I've moved the discussion of well-formedness must go after the
  type system, since we want to separate the preservation vs
  safety challenges
\end{metanote}
\paragraph{Syntax}
Our type system does not type processes directly, but instead focuses on the channels used in the process.
The syntax of \emph{session types} \( S, T \) and \emph{type contexts} \( \Gamma \) is as follows:

\[
  \begin{array}{rcl}
  S,T & ::= & \Tend \mid \Tbase \mid \Tin{S} \mid \Tout{S} \\
    \Gamma &::= & \Cempty \mid \Gamma, l \\
                    \Delta &::= & \Cempty \mid \Cadd{\Delta}{\hastype{x}{S}}
     \end{array}
\]

% \begin{metanote}
%   AM: take the syntax from dual file appst.tex and adapt
% \end{metanote}
% \begin{align*}
%   S,T := &&& \Tend \\
%   |&&& \Tbase \\
%   |&&& \Tin{S} \\
%   |&&& \Tout{S} \\
%   \Gamma := &&& \Cempty \\
%   |&&& \Cadd{\Gamma}{\hastype{x}{S}}
%        %% AM: added
%         \\
%   |&&& \Cadd{\Gamma}{\hastype{l}{\Tbase}} %% changed, AM
% \end{align*}
The \emph{end type}, \( \Tend \), is the type of an endpoint on which no further interaction is possible.
The \emph{base type}, \( \Tbase \), is the type of base values.
The \emph{input type}, \( \Tin{S} \), is the type of an endpoint which is ready to receive a value, then continue with type \( S \).
The \emph{output type}, \( \Tout{S} \), is the type of an endpoint which is ready to send a value, then continue with type \( S \).

Typing contexts gather type information about names and variables. We distinguish between an unrestricted context $\Gamma$ seen as a set: in our special case where we have only one base type, we only record which variable have been used.
%
Type information for channel names is captured in the \emph{linear}
context $\Delta$. We use the comma as split/union, overloaded to
singletons and $\cdot$ is the empty multiset.
% The \emph{empty type context}, \( \Cempty \), does not contain any information.
         %          The \emph{type context with \( x \) added}, \( \Cadd{\Gamma}{\hastype{x}{T}} \) adds the information that \( x \) has type \( T \) to the existing type context \( \Gamma \).
We extend the Barendregt convention to contexts, so that all entries
are distinct.  Note that the order in which information is added to a
type context thus does not matter.
% We treat type contexts as sets, not ordered lists, and the order in which information is added to a type context thus does not matter.

@@

% \begin{metanote}
%   AM: put here the dual rules, remove splitting
% \end{metanote}

% To keep track of linearity, we introduce two \emph{qualifiers}: \( \un{T} \) and \( \lin{T} \).
% A type is \emph{unrestricted}, \( \un{T} \), if and only if \( T = \Tbase \) or \( T = \Tend \).
% All types are \emph{linear}, \( \lin{T} \).
% The qualifiers extend to contexts in the following way: \( q(\Gamma) \) if and only if \( \hastype{x}{T} \in \Gamma \) implies \( q(T) \).

% The \emph{linear split} of a context, \( \Gamma = \Csplit{\Gamma_1}{\Gamma_2} \), is defined by the following rules:
% \begin{mathpar}
%   \inferrule[Split-Empty]{ }{\Cempty = \Csplit{\Cempty}{\Cempty}}
%   \and
%   \inferrule[Split-Un]{\Csplit{\Gamma_1}{\Gamma_2} = \Gamma \\ \un{T}}{\Cadd{\Gamma}{\hastype{x}{T}} = \Csplit{(\Cadd{\Gamma_1}{\hastype{x}{T}})}{(\Cadd{\Gamma_2}{\hastype{x}{T}})}}
%   \and
%   \inferrule[Split-Lin-L]{\Gamma = \Csplit{\Gamma_1}{\Gamma_2}}{\Cadd{\Gamma}{\hastype{x}{S}} = \Csplit{(\Cadd{\Gamma_1}{\hastype{x}{S}})}{\Gamma_2}}
%   \and
%   \inferrule[Split-Lin-R]{\Gamma = \Csplit{\Gamma_2}{\Gamma_2}}{\Cadd{\Gamma}{\hastype{x}{S}} = \Csplit{\Gamma_1}{(\Cadd{\Gamma_2}{\hastype{x}{S}})}}
% \end{mathpar}

@@

We have three typing judgments: one for values, one for names, and one for processes.
The typing rules for values are as follows:
\begin{mathpar}
  \inferrule[T-Base]{ }{\typev{\Gamma}{\hastype{a}{\Tbase}}} \and
  \inferrule[T-Var]{ }{\typev{\Cadd{\Gamma}l }{\hastype{l}{\Tbase}}}
\end{mathpar}
The typing rule for names is, where $\tend \Phi$ denotes that all types in the range of $\Phi$ are $\Tend$:
\begin{mathpar}
  \inferrule[T-Name]{{\tend\Delta }}{\types{{\Gamma; (\Delta,\hastype{x}{T})}}{\hastype{x}{T}}}
\end{mathpar}
The typing rules for processes are as follows:
\begin{mathpar}

  \inferrule[T-Inact]{\tend\Delta }{\types{\Gamma;\Delta}{\Pend}}
  \and
  \inferrule[T-Par]{\types{\Gamma;\Delta_1}{P} \\ \types{\Gamma;\Delta_2}{Q}}
  {\types{\Gamma; \Csplit{\Delta_1}{\Delta_2}}{\Ppar{P}{Q}}}
  \and
  \inferrule[T-Res]{\types{\Gamma; (\Cadd{\Cadd{\Delta}{\hastype{x}{T}}}{\hastype{y}{\dual{T}}}}{P})}{\types{\Gamma}{\Presd{x}{y}{P}}}
  \and
    \inferrule[T-Out]{%\types{\Gamma;\Delta_1}{\hastype{x}{\Tout{T}}} \\
      \typev{\Gamma}{\hastype{v}{\Tbase}} \\ \types{\Gamma; \Cupdate{\Delta}{\hastype{x}{T}}}{P}}{\types{\Gamma; (\Csplit{\Delta}{\hastype{x}{\Tout{T}}})}{\Pout{x}{v}{P}}}
   %  \and
   % \inferrule[T-Out]{\types{\Gamma;\Delta_1}{\hastype{x}{\Tout{T}}} \\
   %   \typev{\Gamma}{\hastype{v}{\Tbase}} \\ \types{\Gamma; (\Cupdate{\Delta}{\hastype{x}{T}})}{P}}{\types{\Gamma; \Csplit{\Delta_1}{\Delta_2}}{\Pout{x}{v}{P}}}
    \and
    \inferrule[T-IN]{%\types{\Gamma;\Delta_1}{\hastype{x}{\Tout{T}}} \\
      %\typev{\Gamma}{\hastype{v}{\Tbase}} \\
      \types{(\Gamma, \hastype l \Tbase); (\Cupdate{\Delta}{\hastype{x}{T}})}{P}}{\types{\Gamma; (\Csplit{\Delta}{\hastype{x}{\Tin{T}}})}{\Pin{x}{l}{P}}}
  % \inferrule[T-IN]{\typev{\Gamma}{\hastype{v}{\Tbase}} \\
  %   \types{(\Gamma, \hastype l \Tbase); \Delta}{\types{\Gamma; \Csplit {\hastype{x}{\Tout{T}}}{\Delta}}
  %     {\Pin{x}{l}{P}}}}
% \and
% \inferrule[UN-L]
% {\types{(\Gamma,l : T); \Delta}{P\{x/ l\}}}
% {\types{\Gamma; (\Delta, x: \un{T})}{P}}
\end{mathpar}

% \begin{mathpar}
%   \inferrule[T-Base]{\un{\Gamma}}{\types{\Gamma}{\hastype{a}{\Tbase}}}
%   \and
%   \inferrule[T-Var]{\un{\Gamma}}{\types{\Cadd{\Gamma}{\hastype{l}{\Tbase}}}{\hastype{l}{\Tbase}}}
% \end{mathpar}
% The only typing rule for names is:
% \begin{mathpar}
%   \inferrule[T-Name]{\un{\Gamma}}{\types{\Cadd{\Gamma}{\hastype{x}{T}}}{\hastype{x}{T}}}
% \end{mathpar}
% The typing rules for processes are as follows:
% \begin{mathpar}
%   \inferrule[T-Inact]{\un{\Gamma}}{\types{\Gamma}{\Pend}}
%   \and
%   \inferrule[T-Par]{\types{\Gamma_1}{P} \\ \types{\Gamma_2}{Q}}{\types{\Csplit{\Gamma_1}{\Gamma_2}}{\Ppar{P}{Q}}}
%   \and
%   \inferrule[T-Res]{\types{\Cadd{\Cadd{\Gamma}{\hastype{x}{T}}}{\hastype{y}{\dual{T}}}}{P}}{\types{\Gamma}{\Presd{x}{y}{P}}}
%   \and
%   \inferrule[T-Out]{\types{\Gamma_1}{\hastype{x}{\Tout{T}}} \\ \types{\Gamma_2}{\hastype{v}{\Tbase}} \\ \types{\Cupdate{\Gamma_3}{\hastype{x}{T}}}{P}}{\types{\Csplit{\Gamma_1}{\Csplit{\Gamma_2}{\Gamma_3}}}{\Pout{x}{v}{P}}}
%   \and
%   \inferrule[T-In]{\types{\Gamma_1}{\hastype{x}{\Tin{U}}} \\ \types{\Cupdate{\Cadd{\Gamma_2}{\hastype{l}{\Tbase}}}{\hastype{x}{U}}}{P}}{\types{\Csplit{\Gamma_1}{\Gamma_2}}{\Pin{x}{l}{P}}}
% \end{mathpar}

\subsubsection{Challenge}
\begin{metanote}
AM:  I'm separating sr from safety, or whatever we call the latter
\end{metanote}
The objective of this challenge is to prove type preservation for our calculus with session types. We start with:
\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  Induction on the reduction relation, TBD.
\end{proof}

The following lemmas may be of use:
\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \) then
    \( \types{\Cadd{\Gamma}{l};\Delta}{P} \).
      \item If \( \types{\Gamma; \Delta}{P} \) and \( \tend{\Delta} \) then
    \( \types{\Gamma;\Cadd{\Delta}{\hastype{x}{T}}}{P} \).
\end{enumerate}
\end{lemma}
\begin{metanote}
  Linear weakening holds only if we put $\tend\Delta$ in T-NAME.
\end{metanote}
\begin{proof}
  By induction on the given derivations.
\end{proof}
\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \( \types{\Gamma; \Delta, z : T}{P} \) and $z\not\in fv(P)$, then \( \types{\Gamma; \Delta}{P} \).
\end{lemma}
\begin{proof}
  By induction on the given derivation.
\end{proof}

\begin{lemma}[Substitution]\mbox{}
  If $\types{(\Gamma, l);\Delta}{P}$ and ${\typev{\Gamma}{\hastype{a}{\Tbase}}}$ then
  \( \types{\Gamma;\Delta}{\subst{P}{l}{a}} \).
\end{lemma}
\begin{proof}
  By induction on the first derivation.
\end{proof}
\begin{lemma}[Preservation for $\equiv$]
  If \( \types{\Gamma;\Delta}{P} \) and \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{metanote}
  AM: I'm still confused about this proof, in particular Vasco's approach to consider the axioms as symmetrization of rewrite rules
\end{metanote}.
\begin{proof}
  By induction on the structure of a process context. In the base case, proceed by case analysis on the \textsl{Sc} rule applied:
  \begin{description}
  \item[Res:] use weakening~\ref{lemma:weak} part 2
  \item[Comm/Assoc:] by rearranging sub-derivations noting that  order does not matter for linear context
  \item[Inact] Left-to-right by
    strengthening~\ref{lemma:strenD}. Vice-versa, by weakening~\ref{lemma:weak} part 2.
  \item[Res-Par] Follows Vasco's proof by case analysis on $x : T$ being linear and applying weak and stren accordingly.
  \item[Res] order does not matter
  \end{description}
  For the step case, TBD
\end{proof}

% \begin{lemma}
%   If \( \types{\Gamma}{P} \) and \( \un{T} \) then \( \types{\Cadd{\Gamma}{\hastype{x}{T}}}{P} \).
% \end{lemma}

% \begin{lemma}
%   Assume \( \types{\Gamma}{P} \) and \( x \notin \freenames{P} \).
%   Then:
%   \begin{enumerate}
%   \item \( \hastype{x}{\Tout{T}} \notin \Gamma \).
%   \item \( \hastype{x}{\Tin{T}} \notin \Gamma \).
%   \item if \( \Gamma = \Cadd{\Gamma'}{\hastype{x}{T}} \) then \( \types{\Gamma'}{P} \).
%   \end{enumerate}
% \end{lemma}

% \begin{lemma}
%   If \( \types{\Gamma}{P} \) and \( \scong{P}{Q} \) then \( \types{\Gamma}{Q} \).
% \end{lemma}
% % \begin{metanote}
% %   AM: the statement of substitution was wrong: now it is consistent with the op sem, but I need to check the (I assume now simpler) proof. Note that $T$ must be \textbf{base} and \( \un{\Gamma_i} \) must also hold.
% % \end{metanote}
% \begin{lemma}
%   If \( \types{\Gamma_1}{\hastype{v}{T}} \) and
%   \( \types{\Cadd{\Gamma_2}{\hastype{l}{T}}}{P} \) and
%   \( \Gamma = \Csplit{\Gamma_1}{\Gamma_2} \) then
%   \( \types{\Gamma}{\subst{P}{l}{v}} \).
% \end{lemma}
